Instructions for developing new pages for the MarineOmics webpage

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
meth - removed duplicate Intro section

The purpose of this tutorial is to provide an example of identifying loci- and gene-level methylation in bisulfite-treated DNA sequence data. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.

e purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. @https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS.

@@

## What to expect
DNA methylation at CpG locus can vary to 0-100%. Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage. Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below. From an taxa perspective it can vary WIDELY on what to expect. The figure below shows the mehthylation ratio of CpGs across the oyster genome. In some marine invertebrates we see ~15% methylation of CpGs (defined as > 75% methylated) across the genome with most CpGs unmethylated. In these taxa methylation usually occurs in gene bodes.

>>>>

The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. @https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS.

@@

## What to expect

DNA methylation at CpG locus can vary to 0-100%. Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage. Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below. From an taxa perspective it can vary WIDELY on what to expect. The figure below shows the mehthylation ratio of CpGs across the oyster genome. In some marine invertebrates we see ~15% methylation of CpGs (defined as > 75% methylated) across the genome with most CpGs unmethylated. In these taxa methylation usually occurs in gene bodes.
